Decomposing labelled proof theory for~intuitionistic~modal~logic

### 2018-05-01 WiL.tex
some small modif

%%% Systems
\newcommand*{\lab}{\mathsf{lab}}
\newcommand*{\IK}{\mathsf{IK}}

@@

\newcommand*{\labels}[2]{{\color{blue}{#1}\:\colon}{#2}}
\newcommand{\SEQ}{\Rightarrow}
\newcommand*{\DD}{\mathcal{D}}
\newcommand*{\rn}[1]  {\ensuremath{\mathsf{#1}}}
\newcommand*{\invr}[1]{#1^\bullet}
\newcommand*{\rel}{R}
\newcommand*{\rlabrn}[2][]  {\rn{#2}_{R#1}}%^\lab}}
\newcommand*{\llabrn}[2][]  {\rn{#2}_{L#1}}%^\lab}}

@@

%% Title information
\title[Decomposing labelled proof theory for intuitionistic modal logic]{Decomposing labelled proof theory\\ for intuitionistic modal logic}         $$[Short Title] is optional;
                                        %% when present, will be used in
                                        %% header instead of Full Title.
%\titlenote{with title note}             %% \titlenote is optional;
                                        %% can be repeated if necessary;
                                        %% contents suppressed with 'anonymous'
%\subtitle{Subtitle}                     %% \subtitle is optional
%\subtitlenote{with subtitle note}       %% \subtitlenote is optional;
                                        %% can be repeated if necessary;
                                        %% contents suppressed with 'anonymous'

@@

%% Author with single affiliation.
\author{Sonia Marin}
%\authornote{with author1 note}          %% \authornote is optional;
                                        %% can be repeated if necessary
%\orcid{nnnn-nnnn-nnnn-nnnn}             %% \orcid is optional
\affiliation{
%  \position{Position1}
%  \department{Department1}              %% \department is recommended
  \institution{IT-Universitetet i K{\o}benhavn}            %% \institution is required
%  \streetaddress{Street1 Address1}
%  \city{City1}
%  \state{State1}
%  \postcode{Post-Code1}
  \country{Denmark}                    %% \country is recommended
}
%\email{first1.last1@inst1.edu}          %% \email is recommended

\author{Marianela Morales}
%\authornote{with author2 note}          %% \authornote is optional;
%% can be repeated if necessary
%\orcid{nnnn-nnnn-nnnn-nnnn}             %% \orcid is optional
\affiliation{
	%  \position{Position2a}
	%  \department{Department2a}             %% \department is recommended
	\institution{Universidad Nacional de C\'ordoba}           %% \institution is required
	%  \streetaddress{Street2a Address2a}
	%  \city{City2a}
	%  \state{State2a}
	%  \postcode{Post-Code2a}
	\country{Argentina}                   %% \country is recommended
}
%\email{first1.last1@inst1.edu}          %% \email is recommended

@@

\sonia{Removed paragraph. Simpson's system can easily be made multi-conclusion in the style of Maehara without the $\le$-atoms.}\lutz{maybe, but getting all the theorems right is not so easy, we had a hard time in the nested variant}

@@

\section{Preliminaries on intuitionistic modal logics}

@@

\marianela{is this enough as an introduction of the axiomatisations to present the theorem 2.5?}

The axiomatisation that is now generally accepted as intuitionistic modal logic denoted by IK was given by Plotkin and Stirling~\cite{Plotkin} and is equivalent to the one proposed by Fischer-Servi ~\cite{Fischer}, and by Ewald~\cite{Ewald} in the case of intuitionistic tense logic. It then was investigated in detail in ~\cite{Simpson}, in which strong arguments are given in favour of this axiomatic definition: it allows for adapting to intuitionistic logic the standard embedding of modal logic into first-order logic, and also provides an extension of the standard Kripke semantics for classical modal logic to the intuitionistic case.

The Kripke semantics for IK was first defined by Fischer-Servi in ~\cite{Fischer}.
This semantics for intuitionistic modal logic combines the Kripke semantics for intuitionistic propositional logic and the one for classical modal logic, using two distinct relations on the set of worlds.

@@

$w \not\Vdash \bot$

@@

We write $w \not\Vdash A$ if it is not the case that $w \Vdash A$, but contrarily to the classical case, we do not have $w \Vdash \neg A$ iff $w \Vdash A$ (since $\neg A$ is defined as $A \vljm \bot$).

From the monotonicity of the valuation function $V$ , we get a monotonicity property for the relation:
\begin{proposition}(Monotonicity) 
For any formula $A$ and for $w, w' \in W$, if $w \le w'$ and $w \Vdash A$, then $w' \Vdash A$.

\end{proposition}

@@

To show completeness wrt.~Simpson's $\lab\IK$ we obtain the proof for the next proposition:\marianela{I do not know if it is a good idea to add this but...}

\begin{proposition}
	\begin{enumerate}
		\item\label{StoM-seq} If there is a proof $\vlderivation{\vlhtr{\DD}{\G, \Left \SEQ \labels{z}{C}}}$ in $\lab\IK$ then there is a proof $\vlderivation{\vlhtr{\DD^\rn{m}}{\G, \Left \SEQ \labels{z}{C}}}$ in $\lab\heartsuit\IK$.

		\item\label{MtoS-seq} If there is a proof $\vlderivation{\vlhtr{\DD}{\G, \Left \SEQ \Right}}$ in $\lab\heartsuit\IK$ then there is a proof $\vlderivation{\vlhtr{\DD^\rn{s}}{\G, \Left \SEQ \labels{z}{C}}}$ in $\lab\IK$.
	\end{enumerate}
\end{proposition}

On the other hand, as part of the proof of Theorem~\ref{thm:simpson-sound-compl}, we obtain the proof for different lemmas such as:

\begin{lemma}\label{lem:weak}\hbox{}\quad
	\begin{enumerate}
		\item 
		If there exists a proof 
		$\vlderivation{\vlhtr{\DD}{\G, \Left \SEQ \Right, \labels{x}{\BOT}}}$ 
		then there exists a proof 
		$\vlderivation{\vlhtr{\DD^{\bm\BOT}}{\G, \Left \SEQ \Right}}$

		\item 
		If there exists a proof 
		$\vlderivation{\vlhtr{\DD}{\G, \Left, \labels{x}{\TOP} \SEQ \Right}}$ 
		then there exists a proof 
		$\vlderivation{\vlhtr{\DD^{\bm\TOP}}{\G, \Left \SEQ \Right}}$

		\item 
		If there exists a proof 
		$\vlderivation{\vlhtr{\DD}{\G, \Left \SEQ \Right}}$ 
		then there exists a proof 
		$\vlderivation{\vlhtr{\DD^{\rn w}}{\G, x \rel y, u \le v, \Left, \labels{z}{A} \SEQ \Right, \labels{w}{B}}}$
	\end{enumerate}
\end{lemma}

\begin{lemma}\label{lem:inv}\hbox{}\quad
	\begin{enumerate}
	\item
	If there exists a proof 
	$\vlderivation{\vlhtr{\DD}{\G, \Left, \labels{x}{A \IMP B} \SEQ \Right}}$ 
	then there exists a proof 
	$\vlderivation{\vlhtr{\DD^{\invr{\llabrn\IMP}}}{\G, \Left, \labels{y}{B} \SEQ \Right}}$
	of the same (or smaller) height, for any label $y$ that does not appear in $\DD$.

	\item
	If there exists a proof 
	$\vlderivation{\vlhtr{\DD}{\G, \Left \SEQ \Right, \labels{x}{A \IMP B}}}$ 
	then there exists a proof 
	$\vlderivation{\vlhtr{\DD^{\invr{\rlabrn\IMP}}}{\G, x \le y, \Left, \labels{y}{A} \SEQ \Right, \labels{y}{B}}}$
	of the same (or smaller) height, for any label $y$ that does not appear in $\DD$.

	\item 
	If there exists a proof 
	$\vlderivation{\vlhtr{\DD}{\G, \Left \SEQ \Right, \labels{x}{\BOX A}}}$ 
	then there exists a proof 
	$\vlderivation{\vlhtr{\DD^{\invr{\rlabrn\BOX}}}{\G, x \le u, u \rel v, \Left \SEQ \Right, \labels{v}{A}}}$
	of the same (or smaller) height, for any label $u$ and $v$ that do not appear in $\DD$.

	\item 
	If there exists a proof 
	$\vlderivation{\vlhtr{\DD}{\G, \Left, \labels{x}{\DIA A} \SEQ \Right}}$ 
	then there exists a proof 
	$\vlderivation{\vlhtr{\DD^{\invr{\llabrn\DIA}}}{\G, x \rel y, \Left, \labels{y}{A} \SEQ \Right}}$
	of the same (or smaller) height, for any label $y$ that does not appear in $\DD$.

	\end{enumerate}
\end{lemma}

>>>>

%%% Systems
\newcommand*{\ax}[1]{\mathsf{#1}}
\newcommand*{\lab}{\mathsf{lab}}
\newcommand*{\IK}{\mathsf{IK}}

@@

%%% Labelled sequents
\newcommand*{\labels}[2]{{\color{blue}{#1}\:\colon}{#2}}
\newcommand{\SEQ}{\Rightarrow}
\newcommand*{\DD}{\mathcal{D}}
\newcommand*{\rn}[1]  {\ensuremath{\mathsf{#1}}}
\newcommand*{\invr}[1]{#1^\bullet}
\newcommand*{\rel}{R}

%%% Labelled rules
\newcommand*{\labrn}[2][]  {\rn{#2}_{#1}}%^{\lab}}}
\newcommand*{\rlabrn}[2][]  {\rn{#2}_{R#1}}%^\lab}}
\newcommand*{\llabrn}[2][]  {\rn{#2}_{L#1}}%^\lab}}
%
%\DeclareFontShape{U}{MnSymbolC}{m}{n}{
%	<-6>  MnSymbolC5
%	<6-7>  MnSymbolC6
%	<7-8>  MnSymbolC7
%	<8-9>  MnSymbolC8
%	<9-10> MnSymbolC9
%	<10-12> MnSymbolC10
%	<12->   MnSymbolC12}{}
%\DeclareFontShape{U}{MnSymbolC}{b}{n}{
%	<-6>  MnSymbolC-Bold5
%	<6-7>  MnSymbolC-Bold6
%	<7-8>  MnSymbolC-Bold7
%	<8-9>  MnSymbolC-Bold8
%	<9-10> MnSymbolC-Bold9
%	<10-12> MnSymbolC-Bold10
%	<12->   MnSymbolC-Bold12}{}
%
%\DeclareSymbolFont{MnSyC}         {U}  {MnSymbolC}{m}{n}
%%
%\DeclareMathSymbol{\diamondplus}{\mathbin}{MnSyC}{124}
%\DeclareMathSymbol{\boxtimes}{\mathbin}{MnSyC}{117}
%
\newcommand*{\brsym}{\mathord{\scalebox{.8}{$\blacksquare$}}}
\newcommand*{\boxbrn}[1]{\rn{\brsym_\rn{#1}}}%^{\lab}}}
\newcommand*{\diasym}{\mathord{\blacklozenge}}
\newcommand*{\diabrn}[1][]{\rn{\diasym_\rn{#1}}}

@@

%% Title information
\title{Decomposing labelled proof theory for~intuitionistic~modal~logic}         
										%%[Short Title] is optional;
                                        %% when present, will be used in
                                        %% header instead of Full Title.
%\titlenote{with title note}             %% \titlenote is optional;
                                        %% can be repeated if necessary;
                                        %% contents suppressed with 'anonymous'
%\subtitle{Subtitle}                     %% \subtitle is optional
%\subtitlenote{with subtitle note}       %% \subtitlenote is optional;
                                        %% can be repeated if necessary;
                                        %% contents suppressed with 'anonymous'

@@

%% Author with single affiliation.
\author{Sonia Marin}
\authornote{Funded by the Qatar National Research Council project MetaCLF nb. xxx. }          %% \authornote is optional;
                                        %% can be repeated if necessary
%\orcid{nnnn-nnnn-nnnn-nnnn}             %% \orcid is optional
\affiliation{
%  \position{Position1}
%  \department{Department1}              %% \department is recommended
  \institution{IT-Universitetet i K{\o}benhavn}            %% \institution is required
%  \streetaddress{Street1 Address1}
%  \city{City1}
%  \state{State1}
%  \postcode{Post-Code1}
  \country{Denmark}                    %% \country is recommended
}
%\email{first1.last1@inst1.edu}          %% \email is recommended

\author{Marianela Morales}
\authornote{with author2 note}          %% \authornote is optional;
%% can be repeated if necessary
%\orcid{nnnn-nnnn-nnnn-nnnn}             %% \orcid is optional
\affiliation{
	%  \position{Position2a}
	%  \department{Department2a}             %% \department is recommended
	\institution{Universidad Nacional de C\'ordoba}           %% \institution is required
	%  \streetaddress{Street2a Address2a}
	%  \city{City2a}
	%  \state{State2a}
	%  \postcode{Post-Code2a}
	\country{Argentina}                   %% \country is recommended
}
%\email{first1.last1@inst1.edu}          %% \email is recommended

@@

\sonia{Removed paragraph. Simpson's system can easily be made multi-conclusion in the style of Maehara without the $\le$-atoms.}
\lutz{maybe, but getting all the theorems right is not so easy, we had a hard time in the nested variant}
\sonia{I did not mean it like that. I mean that the point of our system is not that it is multi-conclusion (it is that it allows us to give systems for some more logics!). This paragraph is confusing as such because it shifts the attention of the reader. I think that we can actually restrict to single conclusion with no difference.}

@@

\section{Intuitionistic modal logics}

@@

\marianela{is this enough as an introduction of the axiomatisations to present the theorem 2.5?}
\sonia{no, we would need the formal definition. But the text is good.}

The axiomatisation that is now generally accepted as intuitionistic modal logic denoted by IK was given by Plotkin and Stirling~\cite{Plotkin} and is equivalent to the one proposed by Fischer-Servi ~\cite{Fischer}, and by Ewald~\cite{Ewald} in the case of intuitionistic tense logic. 
%
It then was investigated in detail in ~\cite{Simpson}, in which strong arguments are given in favour of this axiomatic definition: 
%
it allows for adapting to intuitionistic logic the standard embedding of modal logic into first-order logic, and also provides an extension of the standard Kripke semantics for classical modal logic to the intuitionistic case.

The Kripke semantics for IK was first defined by Fischer-Servi in ~\cite{Fischer}.
%
This semantics for intuitionistic modal logic combines the Kripke semantics for intuitionistic propositional logic and the one for classical modal logic, using two distinct relations on the set of worlds.

@@

$w \not\Vdash \bot$ (i.e.~it is never the case that $w \Vdash \bot$)

@@

%We write $w \not\Vdash A$ if it is not the case that $w \Vdash A$.%, but contrarily to the classical case, we do not have $w \Vdash \neg A$ iff $w \Vdash A$ (since $\neg A$ is defined as $A \vljm \bot$).

%From the monotonicity of the valuation function $V$ , we get a monotonicity property for the relation:
%\begin{proposition}(Monotonicity) 
%For any formula $A$ and for $w, w' \in W$, if $w \le w'$ and $w \Vdash A$, then $w' \Vdash A$.

%\end{proposition}

@@

%To show completeness wrt.~Simpson's $\lab\IK$ we obtain the proof for the next proposition:\marianela{I do not know if it is a good idea to add this but...}
%
%\begin{proposition}
%	\begin{enumerate}
%		\item\label{StoM-seq} If there is a proof $\vlderivation{\vlhtr{\DD}{\G, \Left \SEQ \labels{z}{C}}}$ in $\lab\IK$ then there is a proof $\vlderivation{\vlhtr{\DD^\rn{m}}{\G, \Left \SEQ \labels{z}{C}}}$ in $\lab\heartsuit\IK$.
%		
%		\item\label{MtoS-seq} If there is a proof $\vlderivation{\vlhtr{\DD}{\G, \Left \SEQ \Right}}$ in $\lab\heartsuit\IK$ then there is a proof $\vlderivation{\vlhtr{\DD^\rn{s}}{\G, \Left \SEQ \labels{z}{C}}}$ in $\lab\IK$.
%	\end{enumerate}
%\end{proposition}
%
%\vspace{2mm}
%
%On the other hand, as part of the proof of Theorem~\ref{thm:simpson-sound-compl}, we obtain the proof for different lemmas such as:
%
%\begin{lemma}\label{lem:weak}\hbox{}\quad
%	\begin{enumerate}
%		\item 
%		If there exists a proof 
%		$\vlderivation{\vlhtr{\DD}{\G, \Left \SEQ \Right, \labels{x}{\BOT}}}$ 
%		then there exists a proof 
%		$\vlderivation{\vlhtr{\DD^{\bm\BOT}}{\G, \Left \SEQ \Right}}$
%		
%		\item 
%		If there exists a proof 
%		$\vlderivation{\vlhtr{\DD}{\G, \Left, \labels{x}{\TOP} \SEQ \Right}}$ 
%		then there exists a proof 
%		$\vlderivation{\vlhtr{\DD^{\bm\TOP}}{\G, \Left \SEQ \Right}}$
%		
%		\item 
%		If there exists a proof 
%		$\vlderivation{\vlhtr{\DD}{\G, \Left \SEQ \Right}}$ 
%		then there exists a proof 
%		$\vlderivation{\vlhtr{\DD^{\rn w}}{\G, x \rel y, u \le v, \Left, \labels{z}{A} \SEQ \Right, \labels{w}{B}}}$
%	\end{enumerate}
%\end{lemma}
%
%\begin{lemma}\label{lem:inv}\hbox{}\quad
%	\begin{enumerate}
%	\item
%	If there exists a proof 
%	$\vlderivation{\vlhtr{\DD}{\G, \Left, \labels{x}{A \IMP B} \SEQ \Right}}$ 
%	then there exists a proof 
%	$\vlderivation{\vlhtr{\DD^{\invr{\llabrn\IMP}}}{\G, \Left, \labels{y}{B} \SEQ \Right}}$
%	of the same (or smaller) height, for any label $y$ that does not appear in $\DD$.
%	
%	\item
%	If there exists a proof 
%	$\vlderivation{\vlhtr{\DD}{\G, \Left \SEQ \Right, \labels{x}{A \IMP B}}}$ 
%	then there exists a proof 
%	$\vlderivation{\vlhtr{\DD^{\invr{\rlabrn\IMP}}}{\G, x \le y, \Left, \labels{y}{A} \SEQ \Right, \labels{y}{B}}}$
%	of the same (or smaller) height, for any label $y$ that does not appear in $\DD$.
%	
%	\item 
%	If there exists a proof 
%	$\vlderivation{\vlhtr{\DD}{\G, \Left \SEQ \Right, \labels{x}{\BOX A}}}$ 
%	then there exists a proof 
%	$\vlderivation{\vlhtr{\DD^{\invr{\rlabrn\BOX}}}{\G, x \le u, u \rel v, \Left \SEQ \Right, \labels{v}{A}}}$
%	of the same (or smaller) height, for any label $u$ and $v$ that do not appear in $\DD$.
%	
%	\item 
%	If there exists a proof 
%	$\vlderivation{\vlhtr{\DD}{\G, \Left, \labels{x}{\DIA A} \SEQ \Right}}$ 
%	then there exists a proof 
%	$\vlderivation{\vlhtr{\DD^{\invr{\llabrn\DIA}}}{\G, x \rel y, \Left, \labels{y}{A} \SEQ \Right}}$
%	of the same (or smaller) height, for any label $y$ that does not appear in $\DD$.
%	
%	\end{enumerate}
%\end{lemma}
%

Proof of $\ax{4_{\BOX}} \colon \BOX A \IMP \BOX\BOX A$

$\vlderivation{
	\vlin{\rlabrn\IMP}{}{\labels{x}{\BOX A \IMP \BOX\BOX A}}{	
		\vliq{\rlabrn\BOX}{}{x \le w, \labels{w}{\BOX A} \SEQ \labels{w}{\BOX\BOX A}}{
			\vlin{\rn{F_1}}{}{x \le w, w \le w', \bm{w' \rel v}, \bm{v \le v'}, v' \rel u, \labels{w}{\BOX A} \SEQ \labels{u}{A}}{
				\vlin{\rn{trans}}{}{x \le w, w \le w', w' \rel v, v \le v', v' \rel u, \bm{w' \le t}, \bm{t \rel v'}, \labels{w}{\BOX A} \SEQ \labels{u}{A}}{
					\vlin{\boxbrn{4}}{}{x \le w, w \le w', w' \rel v, v \le v', \bm{v' \rel u}, w' \le t, \bm{t \rel v'}, w \le t \labels{w}{\BOX A} \SEQ \labels{u}{A}}{
						\vlin{\labrn\BOX}{}{x \le w, w \le w', w' \rel v, v \le v', v' \rel u, w' \le t, t \rel v', w \le t, \bm{t \rel u}, \labels{w}{\BOX A} \SEQ \labels{u}{A}}{
							\vlin{\labrn{id}}{}{x \le w, w \le w', w' \rel v, v \le v', v' \rel u, w' \le t, t \rel v', w \le t, t \rel u, \labels{w}{\BOX A}, \labels{u}{A} \SEQ \labels{u}{A}}{
								\vlhy{}
							}
						}
					}
				}
			}
		}
	}
}
$

Proof of $\ax{4_{\DIA}} \colon \DIA\DIA A \IMP \DIA A$:

$\vlderivation{
	\vlin{\rlabrn\IMP}{}{\labels{x}{\DIA\DIA A \IMP \DIA A}}{
		\vliq{\llabrn\DIA}{}{x \le w, \labels{w}{\DIA\DIA A} \SEQ \labels{w}{\DIA A}}{
			\vlin{\diasym_\rn{4}}{}{x \le w, w \rel v, v \rel u, \labels{u}{A} \SEQ \labels{w}{\DIA A}}{
				\vlin{\rlabrn\DIA}{}{x \le w, w \rel v, v \rel u, u \le u', w \rel u'  \labels{u}{A} \SEQ \labels{w}{\DIA A}}{
					\vlin{\labrn{id}}{}{x \le w, w \rel v, v \rel u, u \le u', w \rel u'  \labels{u}{A} \SEQ \labels{w}{\DIA A}, \labels{u'}{A}}{
						\vlhy{}
					}
				}
			}
		}
	}
}
$
